Wigglyhedra

### 2024-06-12 draft/wigglyhedra.tex
improve proof bijection

\vincent{Here, I am using the fact that a set of pairwise non-crossing arcs can be drawn simultaneously non-crossing. I am not sure I want to enter these details.}

@@

\vincent{Asilata has an alternative description of the flip...}

@@

\begin{definition}
A \defn{wiggly permutation} is a permutation of~$[2n]$ which avoids the patterns:
\begin{itemize}
\item $(2j-1) \cdots i \cdots (2j)$ for~$j \in [n]$ and~$i < 2j-1$,
\item $(2j) \cdots k \cdots (2j-1)$ for~$j \in [n]$ and~$k > 2j$.
\end{itemize}
\end{definition}

@@

Recall that the \defn{inversion set} of a permutation~$\sigma$ of~$[2n]$ is the set
\[
\inv(\sigma) \eqdef \set{\big( \sigma(i), \sigma(j) \big)}{1 \le i < j \le 2n \text{ and } \sigma(i) > \sigma(j)}.
\]
We denote by~$\ninv(\sigma)$ the non-inversions of~$\sigma$, that is the complement of~$\inv(\sigma)$.
Note that~$\inv(\sigma)$ and~$\ninv(\sigma)$ are transitive: $\{(k,j), (j,i)\} \subseteq \inv(\sigma)$ implies~$(k,i) \in \inv(\sigma)$ for all~${1 \le i < j < k \le 2n}$, and similarly for~$\ninv$.
The inversion sets of wiggly permutations are characterized as follows.

@@

Note that unfortunately, wiggly permutations do not form a zigzag language in the sense of~\cite{HartungHoangMutzeWilliams} (which would have guarantied the existence of a Gray code).

@@

\begin{definition}
\label{def:bijection1}
Consider a wiggly pseudotriangulation~$T$.
Rotating counterclockwise inside each wiggly pseudotriangle of~$T$, we label by~$2h-1$ (resp.~$2h+1$) the corner immediately preceding (resp.~following) the hinge~$h$ (the remaining corner remains unlabeled).
We define~$\Phi(T)$ as the permutation obtained by reading these labels from bottom to top, meaning that for each internal wiggly arc~$\alpha$ of~$T$, the label incident to~$\alpha$ and below~$\alpha$ appears just before the label incident to~$\alpha$ and above~$\alpha$.
\end{definition}

@@

\begin{proof}
  % We first have to prove that~$\Phi$ and~$\Psi$ are well-defined.
  We break the proof into several steps.

@@

  Finally, one can check using~\cref{lem:wigglyCoverRelation} that increasing flips in triangulations correspond precisely to edges in the wiggly lattice.
  \asilata{Todo: more details here?}
\item

\end{proof}

@@

We now define two families of vectors, illustrated in \cref{fig:pseudotriangulationMatrices}.
%
\begin{figure}
\centerline{\raisebox{-1.8cm}{\includegraphics[scale=2]{pseudotriangulationMatrices}} \quad \input{figures/gcmatrices}}
\caption{The $\b{g}$-matrix and $\b{c}$-matrix of a wiggly pseudotriangulation~$T$. The $i$th column of these matrices are the $\b{g}$-vector~$\b{g}(\alpha)$ and $\b{c}$-vector~$\b{c}(\alpha, T)$ of the $i$th wiggly arc~$\alpha$ of~$T$ (ordered from bottom to top). The colors of the columns match the colors of the wiggly arcs.}
\label{fig:pseudotriangulationMatrices}
\end{figure}

@@

A few additional questions to arouse your curiosity:
\begin{enumerate}
\item Can the construction of \cref{thm:wigglyhedron} be adapted to provide a more combinatorial construction of the polytope of pseudotriangulations~\cite{RoteSantosStreinu-polytope}, that would only depend on the order type (aka oriented matroid~\cite{BjornerLasVergnasSturmfelsWhiteZiegler}) of the point configuration?
\item What is the dual interpretation of wiggly pseudotriangulations as pseudoline arrangements? (see~\cite{PilaudPocchiola} for context) Can this interpretation be extended to other finite Coxeter groups? (see~\cite{CeballosLabbeStump} for context).
\item What is the multi wiggly complex?
\end{enumerate}

@@

Let \(P_i\) denote the indecomposable projective \(A_m\)-module defined as \(P_i = A_m\cdot e_i\).
For any integer~\(j\), let \(P_i \langle j \rangle\) denote its graded shift by \(j\).
There is an embedding of \(A_m\text{-mod}\) into \(D^b(A_m\text{-mod})\), which sends any \(A_m\)-module as a complex of \(A_m\)-modules that is concentrated in degree \(0\).
In particular, the objects \(P_i\langle j \rangle\) can be regarded as objects of \(D^b(A_m\text{-mod})\).
\begin{definition}
  The category \(\mathcal{C}_m\) is defined to be the smallest triangulated subcategory of \(D^b(A_m\text{-mod})\) containing all the objects \(P_i\langle j \rangle\).
\end{definition}
For further discussion on the construction, see, e.g.,~\cite[Sect.~2.3.3]{bap.deo.lic:22}.
The category \(\mathcal{C}_m\) is \(k\)-linear and triangulated, where \([1]\) denotes the triangulated shift.
For \(X, Y \in \mathcal{C}_m\), let \(\Hom^{g,h}(X,Y)\) denote the space of morphisms in \(\mathcal{C}_m\) of internal degree \(g\) and homological degree \(h\).
In other words, we have
\[\Hom^{g,h}(X,Y) = \Hom(X,Y\langle g \rangle[h]).\]
Furthermore, for any integers \(a\) and \(b\), we have
\[\Hom^{g,h}(X\langle g' \rangle[h'], Y\langle g' \rangle[h']) = \Hom^{g,h}(X,Y).\]

@@

We refer the reader to e.g.~\cite[Sect.~1.3]{bel.ber.del:82} for the definition of a \(t\)-structure on a triangulated category.
The following proposition is easily checked using the calculations in~\cref{rem:generating-morphisms} and standard techniques involving categories of complexes; we omit the proof.
\begin{proposition}
  The pair of subcategories \((\mathcal{C}_m^{\leq 0}, \mathcal{C}_m^{\geq 0})\) forms a bounded \(t\)-structure on \(\mathcal{C}_m\).
\end{proposition}
We call this \(t\)-structure the \defn{standard \(t\)-structure} on \(\mathcal{C}_m\).
Set \(\mathcal{A}_m = \mathcal{C}^{\leq 0}_m \cap \mathcal{C}_m^{\geq 0}\); this is the smallest full subcategory of \(\mathcal{C}_m\) closed under isomorphism, which contains all linear complexes.
It follows that \(\mathcal{A}_m\) is abelian.

@@

\subsection{The Khovanov--Seidel curve model}
In this section we briefly recall how to recover an object of \(\mathcal{C}_{n+1}\) from a wiggly arc on \(n+2\) points on a line.
In fact, the procedure works on more general arcs as well --- namely, those that are not necessarily \(x\)-monotone.
All the results of this section are due to Khovanov and Seidel~\cite{kho.sei:02}.

@@

In particular, the line segment joining point \(i\) to point \(i+1\) is sent to the object \(P_{i+1}\) in the category \(\mathcal{C}_{n+1}\).
It is easy to check that the (minimal) intersection numbers between any two such segments counts the number of morphisms\asilata{Discuss (or choose to ignore) grading.} between the associated objects.
In particular, the self-intersection number of a single segment is computed to be \(2\), as the second copy of the segment can be slightly deformed to be disjoint from the first copy except at the two endpoints.
In fact, it is true for any pair of curves that its minimal intersection number is the number of morphisms between the associated objects~\cite[Lem.~4.11]{kho.sei:02}.
\asilata{Add picture of curve cutting.}

@@

\subsection{Categorical constraints on compatibility of curves}
In this section, we give a categorical criterion for two wiggly arcs to be compatible as in~\cref{def:compatible}.
As such, this section is somewhat dual to the previous section --- the criterion in this section determines whether a pair of wiggly arcs can belong together in a single wiggly pseudotriangulation, whereas the previous section generated a wiggly pseudotriangulation given any object of the category.
\begin{proposition}
  Consider two wiggly arcs \(\alpha\) and \(\alpha'\), and let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to \(\alpha\) and \(\alpha'\) respectively.
  Then \(\alpha\) and \(\alpha'\) are compatible in the sense of~\cref{def:compatible} if and only if \(\Ext^1(X, X') = 0\).
\end{proposition}
\begin{proof}
  \asilata{TODO.}
\end{proof}
\begin{remark}
  This proposition is similar to the condition of being compatible in a cluster-tilting object.
  \asilata{Expand and make precise.}
\end{remark}
\begin{remark}
  As stated, this proposition is not expected to be true in greater generality.
  For instance, it does not hold for the zigzag algebra of Dynkin type~$D_4$.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
  \asilata{Decide what to say here, because the forthcoming work is far from complete, although I do have counterexamples in type \(D_4\). Discussing how to interpret these counterexamples may not involve too much more work?}
\end{remark}
\begin{question}
  How to best relate this section to the previous one purely categorically?
  That is, given the categorical constraint on compatibility of arcs, is it also clear that it gives a constraint on objects appearing in the cohomology filtration?
  \asilata{Make more precise.}
\end{question}

\asilata{Write something about \(g\)-vectors.}

>>>>

%\vincent{Here, I am using the fact that a set of pairwise non-crossing arcs can be drawn simultaneously non-crossing. I am not sure I want to enter these details.}

@@

%\vincent{Asilata has an alternative description of the flip...}

@@

\begin{definition}
\label{def:wigglyPermutation}
A \defn{wiggly permutation} is a permutation of~$[2n]$ which avoids the patterns:
\begin{itemize}
\item $(2j-1) \cdots i \cdots (2j)$ for~$j \in [n]$ and~$i < 2j-1$,
\item $(2j) \cdots k \cdots (2j-1)$ for~$j \in [n]$ and~$k > 2j$.
\end{itemize}
\end{definition}

@@

Recall that the \defn{inversion set} and \defn{non-inversion set} of a permutation~$\sigma$ of~$[2n]$ are the sets
\begin{align*}
\inv(\sigma) & \eqdef \set{\big( \sigma(i), \sigma(j) \big)}{1 \le i < j \le 2n \text{ and } \sigma(i) > \sigma(j)}, \\
\ninv(\sigma) & \eqdef \set{\big( \sigma(i), \sigma(j) \big)}{1 \le i < j \le 2n \text{ and } \sigma(i) < \sigma(j)}.
\end{align*}
Note that~$\inv(\sigma)$ and~$\ninv(\sigma)$ are transitive: $\{(c,b), (b,a)\} \subseteq \inv(\sigma)$ implies~$(c,a) \in \inv(\sigma)$ for all~${1 \le a < b < c \le 2n}$, and similarly for~$\ninv$.
The inversion sets of wiggly permutations are characterized as follows.

@@

Note that unfortunately, wiggly permutations do not form a zigzag language in the sense of~\cite{HartungHoangMutzeWilliams} (which would have guarantied the existence of a Gray code).
\vincent{Ask Aaron what he prefers here...}

@@

\begin{definition}
\label{def:bijection1}
Consider a wiggly pseudotriangulation~$T$.
Rotating counterclockwise inside each wiggly pseudotriangle of~$T$, we label by~$2h-1$ (resp.~$2h+1$) the corner immediately preceding (resp.~following) the hinge~$h$ (the remaining corner remains unlabelled).
We define~$\Phi(T)$ as the permutation obtained by reading these labels from bottom to top, meaning that for each internal wiggly arc~$\alpha$ of~$T$, the label incident to~$\alpha$ and below~$\alpha$ appears just before the label incident to~$\alpha$ and above~$\alpha$.
\end{definition}

@@

\begin{proof}
\vincent{I disagree with some points in this proof, so I put an alternative one below...}
  % We first have to prove that~$\Phi$ and~$\Psi$ are well-defined.
  We break the proof into several steps.

@@

  Finally, one can check using~\cref{lem:wigglyCoverRelation} that increasing flips in triangulations correspond precisely to edges in the wiggly lattice.
\end{proof}

\begin{proof}
First, we show that \(\Phi(T)\) is a well-defined wiggly permutation for each wiggly pseudotriangulation \(T\).
Observe first that~\(\Phi(T)\) is by construction a permutation~$\sigma$ of~$[2n]$.
For any \(j \in [n]\), let \(t_j\) be the pseudotriangle of \(T\) that has hinge \(j\).
Suppose that \(2j-1\) appears before \(2j\) in \(\sigma\) for some fixed \(j \in [n]\).
This means that the pseudotriangle~\(t_j\) has its unlabelled corner to the left of its labelled corners.
In other words, its hinge opens to the right (see \eg $t_2$ in~\cref{fig:bijection}).
Consider any label \(i\) that we encounter as we traverse from \(2j-1\) to \(2j\).
This label belongs to some pseudotriangle \(t_h\) with~$h > j$.
In other words, \(\sigma\) avoids the pattern \(2j-1 \cdots i \cdots 2j\) for \(i < 2j-1\).
The other case (in which \(2j\) appears before \(2j-1\) in \(\sigma\)) is symmetric.

Next, we show that \(\Psi(\sigma)\) is a wiggly pseudotriangulation for each wiggly permutation \(\sigma\).
Observe first that for any~$1 \le u < v \le n$, we cannot see~$\{2u, 2v-1\}$ before~$\{2u-1, 2v\}$ in~$\sigma$ as it contains no forbidden pattern of \cref{def:wigglyPermutation}.
Using the notations of \cref{def:bijection2}, this implies that~$i < j$ and that~${]i,j[} = A \sqcup B$, so that~$\alpha(\sigma,k) \eqdef (i, j, A, B)$ is indeed a wiggly arc for each~$k \in [2n-1]$.
By~\cref{prop:wigglyPseudotriangulations}, it thus suffices to check that for each \(k \neq k'\) in \([2n-1]\), the arcs 
\(\alpha(\sigma,k)\) and~\(\alpha(\sigma,k')\) are distinct and compatible.
\begin{enumerate}
\item \textbf{Distinctness.}
Suppose by contradiction that \(\alpha(\sigma,k) = (i,j,A,B) = \alpha(\sigma,k')\) for some~\({k < k'}\).
Let~${\ell \in \sigma([k'] \ssm [k])}$ and~$\bar\ell \eqdef \lceil \ell/2 \rceil$.
As~$\ell \notin \sigma([k])$, we have~$\bar\ell \notin A$.
As~$\ell \in \sigma([k'])$, we have~$\bar\ell \notin B$.
As~$\{2i-1, 2j\} \subseteq \sigma([k])$ and~$\{2i, 2j-1\} \cap \sigma([k']) = \varnothing$ and~$\sigma$ avoids the forbidden pattern of \cref{def:wigglyPermutation}, we have~$2i < \ell < 2j-1$, so that~$i < \bar\ell < j$.
This contradicts the fact that~${]i,j[} = A \sqcup B$.

\item \textbf{Compatibility.}
Let $\alpha(\sigma, k) = (i,j,A,B)$ and $\alpha(\sigma, k') = (i',j',A',B')$ for some \(k < k'\).
It is clear from the definition that \(i \neq j'\), as the sets over which we take the $\max$ and $\min$ to define \(i\) and \(j'\) respectively are disjoint.
Similarly, \(j \neq i'\).
Hence, the two arcs are pointed.
To check that the arcs are non-crossing, we show $(A \cap B') \cup (\{i,j\} \cap B') \cup (A \cap \{i',j'\}) = \varnothing$.
Indeed~$A \cap B' = \varnothing$ since~$A \subseteq A'$ because~$\sigma([k]) \subseteq \sigma([k'])$.
Moreover, \(\{i,j\} \cap B' = \emptyset\) since~\(\{2i - 1, 2j\} \subseteq \sigma([k]) \subseteq \sigma([k'])\).
Finally, \(A \cap \{i',j'\} = \emptyset\) since~\(\{2i', 2j'-1\} \cap \sigma([k]) \subseteq \{2i', 2j'-1\} \cap \sigma([k']) = \varnothing\).
\end{enumerate}

\vincent{I am working here...}
Next, we show that the maps \(\Phi\) and \(\Psi\) are inverse bijections.
Let \(\sigma\) be a wiggly permutation, so that \(\Psi(\sigma)\) is a wiggly pseudotriangulation.
Let~$k \in [2n-1]$ and~$m \eqdef \lceil \sigma(k)/2 \rceil$.
Then observe that
\begin{itemize}
\item if $\{2m, 2m-1\} \subseteq \sigma([k])$, then~$k$ is minimal such that~$\alpha(\sigma,k)$ passes above the point~$m$,
\item if~$\sigma(k) = 2m$ and~$2m-1 \notin \sigma([k])$, then~$k$ is minimal such that~$\alpha(\sigma,k)$ has right endpoint~$m$,
\item if~$\sigma(k) = 2m-1$ and~$2m \notin \sigma([k])$, then~$k$ is minimal such that~$\alpha(\sigma,k)$ has left endpoint~$m$.
\end{itemize}
It follows that in the wiggly pseudotriangulation~\(\Psi(\sigma)\), the arc \(\alpha(\sigma, k)\) is precisely the arc lying immediately above the angle labelled by~\(\sigma(k)\).
Therefore, we have~\(\Phi(\Psi(\sigma)) = \sigma\).

Conversely, let \(T\) be a wiggly pseudotriangulation, so that \(\sigma = \Phi(T)\) is a wiggly permutation.
Then for any \(k \in [2n-1]\), the arc immediately above the angle of~$T$ labelled by~$\sigma(k)$ is precisely the one constructed as~\(\alpha(\sigma, k)\).
This shows that \(\Psi(\Phi(T)) = T\).

Finally, one can check using~\cref{lem:wigglyCoverRelation} that increasing flips in triangulations correspond precisely to edges in the wiggly lattice.
\vincent{Indeed, we need more here. Todo.}
\end{proof}

@@

We now define two families of vectors, illustrated in \cref{fig:pseudotriangulationMatrices}.
\vincent{Move that to the corresponding sections, and delete this section.}
%
\begin{figure}
\centerline{\raisebox{-1.8cm}{\includegraphics[scale=2]{pseudotriangulationMatrices}} \quad \input{figures/gcmatrices}}
\caption{The $\b{g}$-matrix and $\b{c}$-matrix of a wiggly pseudotriangulation~$T$. The $i$th column of these matrices are the $\b{g}$-vector~$\b{g}(\alpha)$ and $\b{c}$-vector~$\b{c}(\alpha, T)$ of the $i$th wiggly arc~$\alpha$ of~$T$ (ordered from bottom to top). The colors of the columns match the colors of the wiggly arcs.}
\label{fig:pseudotriangulationMatrices}
\end{figure}

@@

A few additional questions to arouse your curiosity:
\vincent{Integrate that to the previous section}
\begin{enumerate}
\item Can the construction of \cref{thm:wigglyhedron} be adapted to provide a more combinatorial construction of the polytope of pseudotriangulations~\cite{RoteSantosStreinu-polytope}, that would only depend on the order type (aka oriented matroid~\cite{BjornerLasVergnasSturmfelsWhiteZiegler}) of the point configuration?
\item What is the dual interpretation of wiggly pseudotriangulations as pseudoline arrangements? (see~\cite{PilaudPocchiola} for context) Can this interpretation be extended to other finite Coxeter groups? (see~\cite{CeballosLabbeStump} for context).
\item What is the multi wiggly complex?
\end{enumerate}

@@

Let \(P_i\) denote the indecomposable projective \(A_m\)-module defined as \(P_i = A_m\cdot e_i\).
For any integer~\(j\), let \(P_i \langle j \rangle\) denote its graded shift by \(j\).
There is an embedding of \(A_m\text{-mod}\) into \(D^b(A_m\text{-mod})\), which sends any \(A_m\)-module as a complex of \(A_m\)-modules that is concentrated in degree \(0\).
In particular, the objects \(P_i\langle j \rangle\) can be regarded as objects of \(D^b(A_m\text{-mod})\).
\begin{definition}
  The category \(\mathcal{C}_m\) is defined to be the smallest triangulated subcategory of \(D^b(A_m\text{-mod})\) containing all the objects \(P_i\langle j \rangle\).
\end{definition}
For further discussion on the construction, see \eg~\cite[Sect.~2.3.3]{bap.deo.lic:22}.
The category \(\mathcal{C}_m\) is \(k\)-linear and triangulated, where \([1]\) denotes the triangulated shift.
For \(X, Y \in \mathcal{C}_m\), let \(\Hom^{g,h}(X,Y)\) denote the space of morphisms in \(\mathcal{C}_m\) of internal degree \(g\) and homological degree \(h\).
In other words, we have
\[\Hom^{g,h}(X,Y) = \Hom(X,Y\langle g \rangle[h]).\]
Furthermore, for any integers \(a\) and \(b\), we have
\[\Hom^{g,h}(X\langle g' \rangle[h'], Y\langle g' \rangle[h']) = \Hom^{g,h}(X,Y).\]

@@

We refer the reader to \eg~\cite[Sect.~1.3]{bel.ber.del:82} for the definition of a \(t\)-structure on a triangulated category.
The following proposition is easily checked using the calculations in~\cref{rem:generating-morphisms} and standard techniques involving categories of complexes; we omit the proof.
\begin{proposition}
  The pair of subcategories \((\mathcal{C}_m^{\leq 0}, \mathcal{C}_m^{\geq 0})\) forms a bounded \(t\)-structure on \(\mathcal{C}_m\).
\end{proposition}
We call this \(t\)-structure the \defn{standard \(t\)-structure} on \(\mathcal{C}_m\).
Set \(\mathcal{A}_m = \mathcal{C}^{\leq 0}_m \cap \mathcal{C}_m^{\geq 0}\); this is the smallest full subcategory of \(\mathcal{C}_m\) closed under isomorphism, which contains all linear complexes.
It follows that \(\mathcal{A}_m\) is abelian.

@@

\subsection{The Khovanov--Seidel curve model}
\vincent{Add a picture summarizing this section... Asilata drafts it, I will make it.}
In this section we briefly recall how to recover an object of \(\mathcal{C}_{n+1}\) from a wiggly arc on \(n+2\) points on a line.
In fact, the procedure works on more general arcs as well --- namely, those that are not necessarily \(x\)-monotone.
All the results of this section are due to Khovanov and Seidel~\cite{kho.sei:02}.

@@

In particular, the line segment joining point \(i\) to point \(i+1\) is sent to the object \(P_{i+1}\) in the category \(\mathcal{C}_{n+1}\).
It is easy to check that the (minimal) intersection numbers between any two such segments counts the number of morphisms\asilata{Discuss (or choose to ignore) grading.} between the associated objects.
In particular, the self-intersection number of a single segment is computed to be \(2\), as the second copy of the segment can be slightly deformed to be disjoint from the first copy except at the two endpoints.
In fact, it is true for any pair of curves that its minimal intersection number is the number of morphisms between the associated objects~\cite[Lem.~4.11]{kho.sei:02}.

@@

\subsection{Categorical constraints on compatibility of curves}
In this section, we give a categorical criterion for two wiggly arcs to be compatible as in~\cref{def:compatible}.
As such, this section is somewhat dual to the previous section --- the criterion in this section determines whether a pair of wiggly arcs can belong together in a single wiggly pseudotriangulation, whereas the previous section generated a wiggly pseudotriangulation given any object of the category.
\begin{proposition}
  Consider two wiggly arcs \(\alpha\) and \(\alpha'\), and let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to \(\alpha\) and \(\alpha'\) respectively.
  Then \(\alpha\) and \(\alpha'\) are compatible in the sense of~\cref{def:compatible} if and only if \(\Ext^1(X\langle g \rangle, X\langle g' \rangle) = 0\) for any two integers \(g\) and \(g'\).
\end{proposition}
\begin{proof}
  \asilata{TODO.}
\end{proof}
\begin{remark}
  This proposition is similar to the condition of being compatible in a cluster-tilting object.
  \asilata{Expand and make precise.}
\end{remark}
\begin{remark}
  As stated, this proposition is not expected to be true in greater generality.
  For instance, it does not hold for the zigzag algebra of Dynkin type~$D_4$.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
  \asilata{Decide what to say here, because the forthcoming work is far from complete, although I do have counterexamples in type \(D_4\). Discussing how to interpret these counterexamples may not involve too much more work?}
\end{remark}
\begin{question}
  How to best relate this section to the previous one purely categorically?
  That is, given the categorical constraint on compatibility of arcs, is it also clear that it gives a constraint on objects appearing in the cohomology filtration?
  \asilata{Make more precise.}
\end{question}
